---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Background

This shiny app presents an example of using Bayesian methods to merge expert opinion with observed data improve the predictions.  The example is based on the armadillo hunting patterns of the Ach&eacute; (pronounced Ah-CHAY) Indians.

Investigators studying the Ach&eacute; Indians noted the number of Armadillo killed each day over a 38-day period.  For a given day the number of armadillo killed is represented as $Y_i, i=1,2,\ldots,38$.  As $Y_i$ is a count, a suitable model for the probability that $y$ armadillo are killed is the Poisson distribuion

$$
Y_i|\theta \sim POIS(\theta) \quad i = 1,2,\ldots,38 \mbox{ (iid)}
$$

where $\theta$ is the mean kill rate per day.

Suppose there is interest in predicting the number of armadillo killed on in a given day during a future study.  The investigators believe that a more accurate prediction may be obtained by using Bayesian methods to merge available 'expert' opinion with the data observed in the original study.  

The investigators believe a good prior distribution for $\theta$ is the gamma distibution with pdf

$$
f(x|\alpha,\beta) = \frac{\beta^{\alpha}}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x}, \quad x \in (0, \infty)
$$

where $\alpha$ is the shape parameter and $\beta$ is the rate parameter. The gamma distribution was chosen because it's a flexible distribution and because it's conjugate for Poisson distribution.  

Experts suggest that $\theta \approx 0.5$ and $Pr(\theta \le 2)=0.95$.  Our interest is using this information to determine the values of the parameters $\alpha$ and $\beta$ that reflect our uncertainty about $\theta$ in light of the available expert opinion.  We may interpret the first piece of expert information as meaning that $\mbox{mode}(\theta)=0.5$.  For a gamma distribution (shape-rate paramterization) the equation for the mode is

$$
\frac{\alpha -1}{\beta} = 0.5.
$$

Manipulating this equation results the following expression for the shape parameter

$$
\alpha = 0.5\beta+1.
$$

This expression may be used to determine the value of the shape parameter $\alpha$, such that

$$
Pr\left(\theta \le 2\;|\;0.5\beta+1, \beta\right) = 0.95.
$$

The plot below illustrates the effect of the rate parameter $\beta$ on the value of this probability.  The lines show that $\beta = 2.619$ gives the desired probability.

```{r, echo=FALSE, fig.align='center'}
scale = function(x) return(pgamma(2, shape = 0.5*x+1, rate = x ))

curve(scale(x), 
      xlim = c(0,3),
      lwd = 2,
      ylab = expression(Pr(Theta==2*"|"*beta)),
      xlab = expression(beta),
      axes = F,
      cex.lab = 1.1, font = 2)
box(lwd = 1)
abline(h = 0.95, col = 2, lty = 2)
abline(v = 2.619, col = 2, lty = 2)
axis(side = 1, labels = 2.619, at = 2.619, cex.axis = 1.1)
axis(side = 2, labels = 0.95, at = .95, las = 1, cex.axis = 1.1)
```

With $\beta = 2.619$, we can use the expression for the mode to determine the value of the shape parameter $\alpha$ as

$$
\alpha = 2.619 \cdot 0.5 + 1 = `r 2.619*0.5+1`.
$$

With the parameters determined, we want to make sure that this prior distribution represents our uncertainty about $\theta$.  The plot below shows the pdf of the $GAMMA(2.3095, 2.619)$.

```{r, echo=FALSE, fig.align='center'}
curve(dgamma(x, shape = 2.3095, rate = 2.619),
      lwd = 2,
      las = 1,
      ylab = expression(Pr(Theta==theta)),
      xlab = expression(theta),
      xlim = c(0,4))
rate = 2.619 ; shape = 2.3095
```

The posterior distribution is expressed as

$$
\begin{aligned}
P(\theta|y) &= P(y|\theta)P(\theta)\\\\
&= \prod_{i=1}^{38}\left[\frac{\theta^{y_i}e^{-\theta}}{y_i!}\right]\frac{`r rate`^{`r shape`}}{\Gamma(`r shape`)}\theta^{`r shape`-1}e^{-`r rate`\theta}\\\\
&\propto \prod_{i=1}^{38}\theta^{y_i}e^{-\theta}\theta^{`r shape`-1}e^{-`r rate`\theta}\\\\
&\propto \theta^{\sum_{i=1}^{38}y_i}e^{-38\theta}theta^{`r shape`-1}e^{-`r rate`\theta}\\\\
&\propto \theta^{\sum_{i=1}^{38}y_i+`r shape`-1}e^{-(38+`r rate`)\theta}\\\\
&\propto \theta^{`r 10+shape`-1}e^{-`r 38+rate`\theta}
\end{aligned}.
$$

We see from the above expression that the posterior distribution $P(\theta|y)$ is proportional to the kernel of a $GAMMA(`r 10+shape`,`r 38+rate`)$ distribution.  The plot below shows the difference between the prior and posterior distributions

```{r, echo=FALSE, fig.align='center'}
curve(dgamma(x, shape = 2.3095, rate = 2.619),
      lwd = 2,
      las = 1,
      ylab = expression(Pr(Theta==theta)),
      xlab = expression(theta),
      xlim = c(0,4))
curve(dgamma(x, shape = 10+shape, rate = 38+rate),
      lwd = 2,
      col = 2,
      add = T, n = 300)
legend('topright', bty = 'n', 
       legend = c('Prior', 'Posterior'),
       lwd = 2,
       col = 1:2)
```

